---
title: "R project - cleaning summary"
author: "Jigar Patel"
date: "08/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(patchwork)
```

## Data Cleaning

This part consists of cleaning 3 CSVs.

- `tdf_Stages`
- `stage_date`
- `tdf_winners`

We will go through each of the files and see what columns they consist and what type of data it has.


### `tdf_stages`
```{r}
tdf_stages = read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-07/tdf_stages.csv')

head(x = tdf_stages,2)
```

There are many columns with different data types. First we will check if there are any null values.
```{r}
# making a function to detect null values:
check_null <- function(df){
  as.data.frame(lapply(df, function(x) {length(which(is.na(x)))}))
}

# checking null values for `stage_data`
check_null(tdf_stages)
```

There are no null values, and also the data points seem to be consistent and not vaguely different. Hence cleaning this file is not necessary.

### `stage_data`
```{r}
stage_data1 = read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-07/stage_data.csv')

head(stage_data1,2)
```

#### Checking the null values in the `stage_data`
```{r}
check_null(stage_data1)
```

There are many columns with many null values. We will go through each of the columns and see how we deal with null values.

#### `stage_data$time`
Initial data type = `character` string

Histogram plot (`stage_data$time`)
```{r}
 stage_data <- stage_data1
# Extracting only the numerical part inside the string.
stage_data$time = (str_sub(stage_data$time,start = 1,end=-2))


# Histogram of the numeric time column
hist_num_time <- as.numeric(stage_data$time)
```
```{r,fig.width=4,fig.height=3}
hist(hist_num_time, main = "Histogram of time")
```
$time \sim Unif(a=0,b=60)$ 

The peak at 0 seems to be an error.
We replace the null values with a uniform distribution from 
0 to 60,
```{r}
for (i in 1:nrow(stage_data)){
  if (is.na(stage_data$time[i])){
    stage_data$time[i] = runif(1,min = 0, max = 60)
  }
}

stage_data$time <- as.numeric(stage_data$time)

```
#### `stage_data$age`
```{r,fig.width=7,fig.height=2,warning=F,message=F}
# checking `age` column
mean_age <- mean(na.omit(stage_data$age))
sd_age <- sd(na.omit(stage_data$age))

pl_age_distributions <- 
  ggplot(stage_data) + 
  geom_histogram(aes(x = age),alpha=0.7,position='identity') +
  
  geom_histogram(aes(x = rnorm(nrow(stage_data),
                               mean = mean_age,
                               sd = sd_age)),color='red',
                 alpha=0.3) +
  ggtitle('Age Data along with Normal distribution overlayed (red)')
pl_age_distributions
```
$N(\mu_{age}, \sigma_{age}) \sim \text{Age Data}$ 

Null-values replaced by above distribution.

```{r}
for (i in 1:nrow(stage_data)){
  if (is.na(stage_data$age[i])){
    stage_data$age[i] = rnorm(1,mean = mean_age,sd = sd_age)
    
  }
}

```

#### `stage_data$team`

`NA` or `""` $\implies$ *"No Team"*

Replaced the *empty characters* and *null-values* with *"No Team"*.

#### `stage_data$elapsed`

Convert from `char` data => `int` or `double` data type.


```{r,fig.width=4,fig.height=3}
stage_data$elapsed = (str_sub(stage_data$elapsed,start = 1,end=-2))

# Histogram of the numeric elapsed column
hist(as.numeric(stage_data$elapsed),main = "Histogram of elapsed time",xlab = "Elapsed Time")

for (i in 1:nrow(stage_data)){
  if (is.na(stage_data$elapsed[i])){
    stage_data$elapsed[i] = runif(1,min = 0, max = 60)
  }
}

stage_data$elapsed <- as.numeric(stage_data$elapsed)

```
#### `stage_data$points`

Removing the column due to number of null values.
```{r}
stage_data <- stage_data %>% select(-c("points"))
```

#### `stage_data$bib_number`

```{r,fig.width=5,fig.height=2,warning=F,message=F}
mean_bib <- mean(na.omit(stage_data$bib_number))

pl_bib_number_distributions <-
  ggplot(stage_data) + 
  geom_histogram(aes(x = bib_number)) +
  geom_histogram(aes(rexp(nrow(stage_data),
                          1/mean_bib)),
                 color='red',
                 alpha=0.4) +
  ggtitle("bib_number with Exp distribution (red)")
pl_bib_number_distributions 
```
bib_number $\sim Exp(rate=\dfrac{1}{\mu_{\text{bib_number}}})$

Replacing *null-values* with the values following the above distribution.
```{r}
for (i in 1:nrow(stage_data)){
  if (is.na(stage_data$bib_number[i])){
    stage_data$bib_number[i] = rexp(n = 1,rate = 1/mean_bib)
    
  }
}
```

```{r}
tdf_winners = read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-07/tdf_winners.csv')
```

### `tdf_winners`

Columns consisting of null values:
```{r}
check_null(tdf_winners)
```

#### `tdf_winners$height` & `tdf_winners$weight`

```{r,fig.width=10, fig.height=3,warning=FALSE,message=FALSE}
par(mfrow=c(1,2))

 height_with_distributions <- ggplot(tdf_winners) + geom_histogram(data = tdf_winners,aes(x=tdf_winners$height)) + 
  geom_histogram(data = tdf_winners,
                 aes(x=rnorm(nrow(tdf_winners),
                             mean = mean(na.omit(tdf_winners$height)),
                             sd = sd(na.omit(tdf_winners$height))),
                     color= 'red', alpha=0.4)) + 
  ggtitle("Height with normal distribution (red)")

weight_with_distributions <-ggplot(tdf_winners) + geom_histogram(data = tdf_winners,aes(x=tdf_winners$weight)) + 
  geom_histogram(data = tdf_winners,
                 aes(x=rnorm(nrow(tdf_winners),
                             mean = mean(na.omit(tdf_winners$weight)),
                             sd = sd(na.omit(tdf_winners$weight))),
                     color= 'red', alpha=0.4)) + 
  ggtitle('Weight with normal distribution (red)')

height_with_distributions +
weight_with_distributions

```
Height, Weight $\sim N(\mu_{Height,Weight},\sigma_{Height,Weight})$

Replaced null values with the normally distributed random variables with height's and weight's $\mu$s and $\sigma$s.

```{r}
for (i in 1:nrow(tdf_winners)){
  if (is.na(tdf_winners$height[i])){
    tdf_winners$height[i] = rnorm(1,
                             mean = mean(na.omit(tdf_winners$height)),
                             sd = sd(na.omit(tdf_winners$height)))
    
  }
}

for (i in 1:nrow(tdf_winners)){
  if (is.na(tdf_winners$weight[i])){
    tdf_winners$weight[i] = rnorm(1,
                             mean = mean(na.omit(tdf_winners$weight)),
                             sd = sd(na.omit(tdf_winners$weight)))
    
  }
}
```

#### Removing `nickname`,`full_name`,`died` columns as they are not useful for analysis.

```{r}
tdf_winners <- subset(tdf_winners, select=-c(nickname,full_name,died))
```

#### `tdf_winners$time_overall` & `tdf_winners$time_margin`
```{r,fig.width=7,fig.height=3}
par(mfrow=c(1,2))
hist(tdf_winners$time_overall,main='Histogram of time_overall',xlab = 'time_overall')
hist(tdf_winners$time_margin,main='Histogram of time_margin',xlab='time_margin')
```
Substituting *null-values* with exponentially distributed random variables since they both approximately follow exponential distributions. 

```{r}
# We can substitute the null values with exponential distributed values
for (i in 1:nrow(tdf_winners)){
  if (is.na(tdf_winners$time_overall[i])){
    tdf_winners$time_overall[i]=rexp(1,
                                    1/mean(na.omit(tdf_winners$time_overall)))
  }
}

for (i in 1:nrow(tdf_winners)){
  if (is.na(tdf_winners$time_margin[i])){
    tdf_winners$time_margin[i]=rexp(1,
                                    1/mean(na.omit(tdf_winners$time_margin)))
  }
}

```

All three data-sets now have no-null values. 

Adding new columns will be included in the *Analysis section* for calculating and analyzing new information.

---
